\Large Inferential Analysis of the Supreme Court Citation Network

### 2018-03-20 Tex/paper.tex

Let $C$ be the case-to-case adjacency matrix such that $C_{ij} \in \{0,1\}$ is a binary indicator of whether case $i$ cites $j$. Furthermore, let
$$\mathcal{C} =\{C \in \{0,1\}^{(N \cdot (N-1))/2}: C_{ij} \in \{0,1\} \}$$ 
be the set of all possible adjacency matrices among $N$ cases. Note that the cardinality of $\mathcal{C}$ increases exponentially for every newly added case. The probability function of the c-TERGM is defined as
\begin{equation}
P_{\theta}(C)=\cfrac{\exp\big(\theta^T\cdot h\big(C)\big)\big)}{\sum_{Z^*\in \mathcal{C}}\exp\big(\theta^T\cdot h\big(Z^*)\big)\big)}
\label{ercm}
\end{equation}
where $\theta \in \mathbb{R}^q$ is a $q$-dimensional vector of parameters,  $h: \mathcal{C} \to \mathbb{R}^q$ is a q-dimensional vector of different statistics and $\kappa(\theta) := \sum_{Z^*\in \mathcal{C}}\exp\big(\theta^T\cdot h\big(Z^*)\big)$ is a normalization constant that ensures that Equation (\ref{ercm}) defines a probability function on $\mathcal{C}$.

% Network Statistics
The c-TERGM is specified through 
the decision regarding which network statistics $h(\cdot)$ to incorporate. We include the following statistics for the Supreme Court citation network, which are derived from the interdependence hypotheses described in the previous section:
$$h_{edges}:\mathcal{C} \to \mathbb{R}~~~, ~~~C \to \sum_{ij} C_{ij}, $$
the number of citations.  $h_{edges}$ performs the function of an intercept, and models the expected value of any given edge.
$$h_{outstar}:\mathcal{C} \to \mathbb{R}~~~, ~~~C \to \sum_{i=1}^N {\sum_{j \neq i} c_{ij} \choose 2} $$
the number of out-two-stars. An out-two-star is a configuration in which one node sends to two other nodes. The number of these configurations grows quadratically as the origins of ties concentrate on a few highly active/social senders. The out-two-stars configuration is commonly used to model activity (sender activation) in ERGMs.
$$h_{instar}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C \to \sum_{j=1}^N {\sum_{i \neq j} c_{ij} \choose 2} $$
the number of in-two-stars. An in-two-star is a configuration in which one node receives ties from two other nodes. The number of these configurations grows quadratically as the destinations of ties concentrate on a few highly popular recipients. The out-two-stars configuration is commonly used to model popularity in ERGMs.
\begin{align*}
h_{triangle}:\mathcal{C} \to \mathbb{R}~~~, ~~~C \to & \sum_{j<i<k} C_{ij}\cdot C_{jk} \cdot C_{ki} + C_{ji}\cdot C_{jk} \cdot C_{ki} + C_{ij}\cdot C_{kj} \cdot C_{ki} \\ & + C_{ji}\cdot C_{kj} \cdot C_{ki} + C_{ij}\cdot C_{jk} \cdot C_{ik} + C_{ji}\cdot C_{jk} \cdot C_{ik}  \\ & + C_{ij}\cdot C_{kj} \cdot C_{ik} + C_{ji}\cdot C_{kj} \cdot C_{ik},  
\end{align*}
The number of triangles in the network.  Triangles measure triad closure, as discussed above. $$h_{covariate}:\mathcal{C} \to \mathbb{R}~~~, ~~~C \to \sum_{ij} C_{ij}X_{ij}, $$ the effect of an exogenous covariate ($X$). We include several exogenous covariates based on this standard statistic formulation. These covariates are discussed below.

The structure of this model is very similar to a conventional (T)ERGM setup. The main difference is that  $\mathcal{C}$ excludes adjacency matrices in which there are loops that include edges sent at different time points. Loops can exist in the Supreme Court citation network, but only among cases decided in the same term. The normalizing constant in the denominator of the probability function (i.e., Equation \ref{ercm}) is computationally intractable. This means that straightforward methods of maximum likelihood estimation (MLE) are infeasible with ERGM family models. Methods of Monte Carlo approximation are well established \citep{hunter2006inference,van2009framework,hummel2012improving}, but with over 8,000 vertices, even the methods of Monte Carlo MLE are computationally expensive \citep{schmid2017exponential}. To estimate the parameters of the c-TERGM, we rely on two features of the model and data that facilitate the application of an approximate inference method that is faster than Monte Carlo approximation---maximum pseudolikelihood. First, our sample size is very large. Second, under the c-TERGM we assume that citations are formed in each time period conditional upon the citations that exist in the past.  We estimate the c-TERGM via bootstrapped pseudolikelihood \citep{desmarais2012statistical,desmarais2010consistent}.

>>>>

Let $C_t$ be the case-to-case adjacency matrix at time $t$ such that $C_t^{ij} \in \{0,1\}$ is a binary indicator of whether case $i$ cites $j$. Furthermore, let
$$\mathcal{C}_t =\{C_{\leq t} \in \{0,1\}^{(N_t \cdot (N_t-1))/2}: C_{\leq t}^{ij} \in \{0,1\} \}$$ 
be the set of all possible adjacency matrices among $N_t$ cases. $N_t$ denotes the the number of cases in the network at time $t$. Note that the cardinality of $\mathcal{C}_t$ increases exponentially for every newly added case. The probability of observing $C_t$ given past citations $C_{<t}$, where $C_{\leq t}$ is the network up to time $t$ is defined as
\begin{equation}
P(C_t | C_{<t}, \theta)=\cfrac{\exp\big(\theta^T\cdot h\big(C_{\leq t})\big)\big)}{\sum_{Z_t^*\in \mathcal{C}_t}\exp\big(\theta^T\cdot h\big(Z^*_{\leq t})\big)\big)}
\label{ercm}
\end{equation}
where $\theta \in \mathbb{R}^q$ is a $q$-dimensional vector of parameters,  $h: \mathcal{C}_t \to \mathbb{R}^q$ is a q-dimensional vector of different statistics and $\kappa(\theta) := \sum_{Z^*_t\in \mathcal{C}_t}\exp\big(\theta^T\cdot h\big(Z^*_{\leq t})\big)$ is a normalization constant that ensures that Equation (\ref{ercm}) defines a probability function on $\mathcal{C}_t$.

% Network Statistics
The c-TERGM is specified through 
the decision regarding which network statistics $h(\cdot)$ to incorporate. We include the following statistics for the Supreme Court citation network, which are derived from the interdependence hypotheses described in the previous section:
$$h_{edges}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to \sum_{ij} C_{\leq t}^{ij}, $$
the number of citations.  $h_{edges}$ performs the function of an intercept, and models the expected value of any given edge.
$$h_{outstar}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to \sum_{i=1} {\sum_{j \neq i} C_{\leq t}^{ij} \choose 2} $$
the number of out-two-stars. An out-two-star is a configuration in which one node sends to two other nodes. The number of these configurations grows quadratically as the origins of ties concentrate on a few highly active/social senders. The out-two-stars configuration is commonly used to model activity (sender activation) in ERGMs.
$$h_{instar}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to \sum_{j=1} {\sum_{i \neq j} C_{\leq t}^{ij} \choose 2} $$
the number of in-two-stars. An in-two-star is a configuration in which one node receives ties from two other nodes. The number of these configurations grows quadratically as the destinations of ties concentrate on a few highly popular recipients. The out-two-stars configuration is commonly used to model popularity in ERGMs.
\begin{align*}
h_{triangle}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to & \sum_{j<i<k} C^{ij}\cdot C^{jk} \cdot C^{ki} + C^{ji}\cdot C^{jk} \cdot C^{ki} + C^{ij}\cdot C^{kj} \cdot C^{ki} \\ & + C^{ji}\cdot C^{kj} \cdot C^{ki} + C^{ij}\cdot C^{jk} \cdot C^{ik} + C^{ji}\cdot C^{jk} \cdot C^{ik}  \\ & + C^{ij}\cdot C^{kj} \cdot C^{ik} + C^{ji}\cdot C^{kj} \cdot C^{ik},  
\end{align*}
The number of triangles in the network.  Triangles measure triad closure, as discussed above. $$h_{covariate}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to \sum_{ij} C_{\leq t}^{ij}X_{ij}, $$ the effect of an exogenous covariate ($X$). We include several exogenous covariates based on this standard statistic formulation. These covariates are discussed below.

The structure of this model is very similar to a conventional (T)ERGM setup. The main difference is that  $\mathcal{C}_t$ excludes adjacency matrices in which there are loops that include edges sent at different time points. Loops can exist in the Supreme Court citation network, but only among cases decided in the same term. The normalizing constant in the denominator of the probability function (i.e., Equation \ref{ercm}) is computationally intractable. This means that straightforward methods of maximum likelihood estimation (MLE) are infeasible with ERGM family models. Methods of Monte Carlo approximation are well established \citep{hunter2006inference,van2009framework,hummel2012improving}, but with almost $10,000$ vertices, even the methods of Monte Carlo MLE are computationally expensive \citep{schmid2017exponential}. To estimate the parameters of the c-TERGM, we rely on two features of the model and data that facilitate the application of an approximate inference method that is faster than Monte Carlo approximation---maximum pseudolikelihood. First, our sample size is very large. Second, under the c-TERGM we assume that citations are formed in each time period conditional upon the citations that exist in the past.  We estimate the c-TERGM via bootstrapped pseudolikelihood \citep{desmarais2012statistical,desmarais2010consistent}.
